Solving the maximum independant set problem by generic programming einsum networks

### 2021-06-15 paper/paper.tex
new proof

\newtheorem{theorem}{\textit{Branching Rule}}
\theoremstyle{definition}\newtheorem{definition}{\textit{Definition}}

@@

\section{Automated branching}
% check http://www.tcs.rwth-aachen.de/independentset/ for more rules
So far, we have used the language of einsum networks for contraction.
When using sparse tropical tensor networks to find the maximum independent set, we can introduce a new rule to compress the tensor by removing elements that are not helpful.
As show in \Fig{fig:compressrule}, after we contract the tensors in a subregion $R \subseteq G$ of a graph $G$, and obtain a resulting tensor $A$ of rank $|C|$, where $C$ is the set of vertice tensors at the cut.
Each tensor entry $A_{\sigma}$ defines a local maximum independant set size with a fixed boundary configuration $\sigma \in \{0,1\}^{|C|}$ by marginalizing the inner degrees of freedom.
We call this properly the \textit{local maximum rule}.

@@

@@

We are going the verify the Lemmas used for branching in book~\cite{Fomin2013}.
\begin{theorem}\label{rule:one} % basic
  If a vertex $v$ is in an independent set $I$, then none of its neighbors can be in $I$.
On the other hand, if $I$ is a maximum (and thus maximal) independent set,
and thus if $v$ is not in $I$ then at least one of its neighbors is in $I$.
\end{theorem}

@@

\begin{theorem} % 2.5
Let $G=(V,E)$ be a graph, let $v$ and $w$ be adjacent vertices of $G$ such that $N[v] \subseteq N[w]$. Then
\begin{equation}
\alpha(G)=\alpha(G\backslash w).
\end{equation}
\end{theorem}

Contract $N[w]$, both $\{v, w\}$ disapear from the tensor indices because they are inner degrees of freedom.
If $w$ is one, then $N[v]$ are all zeros, the resulting tensor element can not be larger than setting $v=1$ and $w=0$.
By the maximization rule, the local tensor does not change if we remove $w$.

\begin{theorem} % 2.6
  Let $G = (V, E)$ be a graph and let $v$ be a vertex of $G$. If no maximum
independent set of $G$ contains $v$ then every maximum independent set of $G$ contains
at least two vertices of $N(v)$.
\end{theorem}
Contract $N[v]$, the minimum tensor element is $2$, otherwise letting $v$ be in the independent set is one of the solutions. This is again captured by the \textit{local maximum rule}.

\begin{theorem} % 2.7
Let $G = (V, E)$ be a graph and $v$ a vertex of $G$. Then
\begin{equation}
\alpha(G) = \max(1 + \alpha(G \backslash N[v]), \alpha(G \backslash (M(v) \cup \{v\})).
\end{equation}
\end{theorem}

Here, $M(v)$ is the set of mirrors of $v$ in $G$.
A vertex $w \in N^2(v)$ is called a mirror of $v$ if $N(v) \backslash N(w)$ is a clique.
This rule states that if $v$ is not in $M$, there exists an MIS $I$ that $M(v)\notin I$.
otherwise, there must be one of $N(v)$ in the MIS (\textit{local maximum rule}).
If $w$ is in $I$, then none of $N(v) \cap N(w)$ is in $I$, then there must be one of node in the clique $N(v)\backslash N(w)$ in $I$ (\textit{local maximum rule}),
since clique has at most one node in the MIS, the tensor compression will eliminate this solution by moving the occuppied node to the interior.
Hence, the \textit{least restrictive rule} captures the mirror rule.

\begin{theorem} % 2.9
Let $G = (V, E)$ be a graph and $v$ be a vertex of $G$ such that $N[v]$ is a
clique. Then
\begin{equation}
\alpha(G) = 1 + \alpha(G \backslash N[v]).
\end{equation}
\end{theorem}

Contract $N[v]$, and this rule can be captured by the \textit{local maximum rule}.

\begin{theorem}  %2.10
Let $G$ be a graph, let $S$ be a separator of $G$ and let $I(S)$ be the set of
all subsets of $S$ being an independent set of $G$. Then
\begin{equation}
\alpha(G) = \max_{A\in I(S)} |A| + \alpha(G \backslash (S \cup N[A])).
\end{equation}
\end{theorem}
This rule corresponds to first contract tensors in subgraph $S$, then $N[S]$, and then contract the rest parts.
These branching rule can be captured by the \textit{local maximum rule}.

\begin{theorem}  %2.11
Let $G = (V, E)$ be a disconnected graph and $C \subseteq V$ a connected component of $G$. Then
\begin{equation}
\alpha(G) = \alpha(G[C]) + \alpha(G \backslash C)).
\end{equation}
\end{theorem}
Contract by the disconnected parts, and this rule can be captured by the \textit{local maximum rule}.

>>>>

\newtheorem{theorem}{\textit{Theorem}}
\newtheorem{corollary}{\textit Branching Rule}
\theoremstyle{definition}\newtheorem{definition}{\textit{Definition}}

@@

\section{Automated branching}
% check http://www.tcs.rwth-aachen.de/independentset/ for more rules
So far, we have used the language of einsum networks for contraction.
When using sparse tropical tensor networks to find the maximum independent set, we can introduce a new rule to compress the tensor by removing elements that are not helpful.
As show in \Fig{fig:compressrule}, after we contract the tensors in a subregion $R \subseteq G$ of a graph $G$, and obtain a resulting tensor $A$ of rank $|C|$, where $C$ is the set of boundary vertices defined as $C := \{c | c\in R \land c \in G\backslash R\}$.

We denote the effective branching number of contracting the local degrees of freedoms as $|T_S|_{\neq -\infty}/2^{|S|}$, where $|\cdot|_{\neq -\infty}$ is the number of nonzero elements in a tropical tensor.

Each tensor entry $A_{\sigma}$ defines a local maximum independant set size with a fixed boundary configuration $\sigma \in \{0,1\}^{|C|}$ by marginalizing the inner degrees of freedom.
We call this properly the \textit{local maximum rule}.

@@

\begin{theorem}
    A MIS-compact tropical tensor is optimal, i.e. none of its none zero entries can be further reduces.
\end{theorem}

\begin{proof}
    Let us prove it by showing any nonzero entry in a MIS-compact tropical tensor can be the only solution of maximum independent set under a certan exterior graph configuration.
    Let $A$ be a tropical tensor, and an entry of it being $A_{\sigma} = k$, where $\sigma$ is the bounary configuration.
    Construct a graph $G$ such that for a $v \in C$, if $\sigma_v=1$, $\alpha(N[v] \cap (G \backslash R)) = 0$, otherwise, $\alpha(N[v] \cap (G\backslash R)) = \infty$.
    The simplest construction is connecting vertices that $\sigma_v=0$ with infinite many mutually disconnected vertices as illustrated in the following graph.

    \centerline{\includegraphics[width=0.35\textwidth, trim={0cm 2cm 6cm 0cm}, clip]{proofoptimal.pdf}}

    Then we have $\alpha(G) = \infty (|C|-|\sigma|) + k$. If there exists another configuration $\tau$ that generating the same good result, then $\tau \prec \sigma$,
    otherwise it will suffer from infinite punishment from $G\backslash R$.
    Since this tropical tensor is MIS-compact, it is not possible to find a such $\tau$ that $A_\tau \geq k$, otherwise $\sigma$ will be remove by the \textit{least restrictive rule}.
\end{proof}

@@

\begin{corollary}\label{rule:one} % basic
  If a vertex $v$ is in an independent set $I$, then none of its neighbors can be in $I$.
On the other hand, if $I$ is a maximum (and thus maximal) independent set,
and thus if $v$ is not in $I$ then at least one of its neighbors is in $I$.
\end{corollary}

@@

\centerline{\includegraphics[width=0.4\columnwidth,trim={0 3.5cm 0 1cm},clip]{../notebooks/basic.pdf}}

After contracting $N[v]$, $v$ becomes an internal degree of freedom.
Applying tensor compression rule \Eq{eq:compress}, the resulting rank 4 tropical tensor is

\begin{align}
    T_{juwk} = \left(\begin{matrix}
        \left(\begin{matrix}
        ~~~~1 & -\infty \\
        -\infty & ~~~~2
        \end{matrix}\right)_{ju}&
        \left(\begin{matrix}
        -\infty & ~~~~2 \\
        ~~~~2 & ~~~~3
        \end{matrix}\right)_{ju}\\
        \left(\begin{matrix}
        -\infty & ~~~~2 \\
        ~~~~2 & ~~~~3
        \end{matrix}\right)_{ju} &
        \left(\begin{matrix}
        ~~~~2 & ~~~~3 \\
        ~~~~3 & ~~~~4
        \end{matrix}\right)_{ju}
    \end{matrix}\right)_{wk}.
\end{align}

The effective branching value is $11^{1/5} \approx 1.6154$, which is larger than the branching number $\tau(1, 5) \approx 1.3247$.
It does not mean the tropical tensor does not find all the branches, if we contract $N^2[v]$.

\begin{corollary}[mirror rule] % 2.7
For some $v \in V$, a node $u \in N^2(v)$ is called mirror of $v$, if $N(v) \backslash N(u)$ is a clique. We denote the set of of a node $v$ mirrors~\cite{Fomin2013} by $M(v)$.
Let $G = (V, E)$ be a graph and $v$ a vertex of $G$. Then
\begin{equation}
\alpha(G) = \max(1 + \alpha(G \backslash N[v]), \alpha(G \backslash (M(v) \cup \{v\})).
\end{equation}
\end{corollary}

This rule states that if $v$ is not in $M$, there exists an MIS $I$ that $M(v)\notin I$.
otherwise, there must be one of $N(v)$ in the MIS (\textit{local maximum rule}).
If $w$ is in $I$, then none of $N(v) \cap N(w)$ is in $I$, then there must be one of node in the clique $N(v)\backslash N(w)$ in $I$ (\textit{local maximum rule}),
since clique has at most one node in the MIS, the tensor compression will eliminate this solution by moving the occuppied node to the interior.
Hence, the \textit{least restrictive rule} captures the mirror rule.

In the following example, since $u\in N^2(v)$ and $N(v) \backslash N(u)$ is a clique, $u$ is a mirror of $v$.

\centerline{\includegraphics[width=0.4\columnwidth,trim={0 3.5cm 0 1cm},clip]{../notebooks/mirror.pdf}}

After contracting $N[v]\cup u$, $v$ becomes an internal degree of freedom.
Applying tensor compression rule \Eq{eq:compress}, the resulting rank 4 tropical tensor is

\begin{align}
    T_{juwk} = \left(\begin{matrix}
        \left(\begin{matrix}
        ~~~~1 & ~~~~2 \\
        -\infty & -\infty
        \end{matrix}\right)_{ju}&
        \left(\begin{matrix}
        -\infty & -\infty \\
        ~~~~2 & -\infty
        \end{matrix}\right)_{ju}\\
        \left(\begin{matrix}
        -\infty & -\infty \\
        -\infty & -\infty
        \end{matrix}\right)_{ju} &
        \left(\begin{matrix}
        -\infty & -\infty \\
        -\infty & -\infty
        \end{matrix}\right)_{ju}
    \end{matrix}\right)_{wk}.
\end{align}

In this case, the effective branching number is $3^{1/5}\approx 1.2457$,
which is smaller than the branching number $\tau(4, 2) = 1.2721$ by simply applying the mirror rule.

\begin{corollary}[satellite rule] % satellite rule
Let $G$ be agraph $v \in V$. A node $u \in N^2(v)$ is called satellite~\cite{Kneis2009} of $v$, if there is some $u' \in N(v)$ such that $N[u'] \backslash N[v] = \{u\}$.
The set of satellites of a node $v$ is denotedby $S(v)$, and we also use the notation $S[v] := S(v) \cup {v}$. Then 
\begin{equation}
\alpha(G) = \max\{\alpha(G \backslash \{v\}), \alpha(G \backslash N[S[v]]) + |S(v)| + 1\}.
\end{equation}
\end{corollary}

This rule can be capture by contracting $N[v] \cup S(v)$.
In the following example, since $u \in N^2(v)$ and $w \in N(v)$ satisfies $N[w] \backslash N[v] = \{u\}$, $u$ is a satellite of $v$.

\centerline{\includegraphics[width=0.4\columnwidth,trim={0 3.5cm 0 1cm},clip]{../notebooks/satellite.pdf}}

After contracting $N[v] \cup u$, both $v$ and $w$ become internal degrees of freedoms.
Applying tensor compression rule \Eq{eq:compress}, the resulting rank 3 tropical tensor is
\begin{align}
    T_{juk} = \left(\begin{matrix}
        \left(\begin{matrix}
        ~~~~1 & ~~~~2 \\
        ~~~~2 & -\infty
        \end{matrix}\right)_{ju}\\
        \left(\begin{matrix}
        -\infty & -\infty \\
        -\infty & -\infty
        \end{matrix}\right)_{ju}
    \end{matrix}\right)_{k}.
\end{align}

There are 3 nonzero entries. The internal configurations of entry $T(j=1, u=0, k=0) = 2$ is $(v=0, w =1)$,
that of entry $T(j=0, u=1, k=0)=2$ is $(v=1, w=0)$, and that of entry $T(j=0, u=0, k=0)=1$ is $(v=1, w=0)$ or $(v=0, w=1)$.
For entry $T(j=0, u=0, k=0)=1$, we post-select the internal degree of freedom as $(v=0, w=1)$.
Then we can see the satellite rule either ${v, u} \in I$ or $v \notin I$ is satisfied.
In this case, the effective branching number is $3^{1/5}\approx 1.2457$.
